Ecological management of stochastic systems with long transients

### 2020-06-22 manuscript/manuscript.Rmd
WIP

\begin{equation}
X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{logistic growth}}  - 
\underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{saturating consumption}} + 
\underbrace{\xi_t X_t}_{\textrm{environmental noise}} \label{model}
\end{equation}

@@

@@

Under appropriate choice of parameters, the model is bistable, with a low, "endemic" stable state, and a much higher stable "outbreak" state.  In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emerging from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses. This saturating response and the resulting threshold dynamics in the response of affected trees exhibited by this model is quite well motivated   [e.g. @Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate. Initially this results in the strengthening (concave up) portion of the S curve as other defenses are induced by the attack.  However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defenses, leading to a saturating response.  Low beetle densities are thus stable, kept in check by the increasingly strong response of healthy trees.  Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defenses of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting available tree habitat $K$ [@Harvey2014]. Thus our simple dynamical model from Eq 1 captures the qualitative processes and dynamics expected in the beetle outbreak system. Bear in mind that this model is not intended to be a precise, quantitative match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Fig 2. Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B). Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.  Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastic escape from a stable node.

@@

$$\max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)$$
where the evolution of $X_t$ is determined by Eq \eqref{model}. Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.  By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985, @Marescot2013].

>>>>

\begin{equation}
X_{t+1} = g(X_t) = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{logistic growth}}  - 
\underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{saturating consumption}} + 
\underbrace{\xi_t X_t}_{\textrm{environmental noise}} \label{model}
\end{equation}

@@

We can extend this model to include the application of a control action in year $t$ of $A_t$ that decreases the population by a fixed amount before reproduction occurs, such that:

\begin{equation}
X_{t+1} = f(X_t, A_t) = g(X_t - A_t)
\end{equation}

Thus with $A_t = 0$ we can model the dynamics without control, and can model any control policy over time as a set of actions $A_t$ describing the beetle mortality imposed by management response that year.

@@

Under appropriate choice of parameters, the model is bistable, with a low, "endemic" stable state, and a much higher stable "outbreak" state.  Population growth is subject to both the intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses. $t+1$ denotes the species generation time, which may be considerably shorter than an annual cycle.  This saturating response and the resulting threshold dynamics in the response of affected trees exhibited by this model is quite well motivated [e.g. @Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate. Initially this results in the strengthening (concave up) portion of the S curve as other defenses are induced by the attack.  However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defenses, leading to a saturating response.  Low beetle densities are thus stable, kept in check by the increasingly strong response of healthy trees.  Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defenses of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting available tree habitat $K$ [@Harvey2014]. Thus our simple dynamical model from Eq 1 captures the qualitative processes and dynamics expected in the beetle outbreak system. Bear in mind that this model is not intended to be a precise, quantitative match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].  A decrease in the parameter $a$ could be thought to model the weakened ability of the trees to respond to attack, arising from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

This same model can thus produce two very different realizations depending on the timing of the outbreak.  It is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastic escape from a stable node, as illustrated in the ensemble simulations shown in Fig 2. Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B). Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.

@@

\begin{equation}
V_t(X_t) = \max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)
\end{equation}

where the evolution of $X_t$ is determined by Eq \eqref{model}. Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.  By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985; @Marescot2013].
